On the reality of the quantum state once again

### 2021-12-31 Articles/2021-HarriganSpekkens/HarriganSpekkens.tex
HarriganSpekkens: section 2 and figures added

\documentclass[10pt,twocolumn, nofootinbib]{revtex4-2}
%\documentclass[aps,pra,10pt,twocolumn,floatfix,nofootinbib]{revtex4-1}
%\documentclass[10pt,twocolumn,letterpaper]{article}

\RequirePackage{amsthm}
\RequirePackage{amsmath}
\RequirePackage{amssymb}
\RequirePackage{enumitem}
\setlist{noitemsep}
\usepackage{graphicx}
\usepackage{hyperref}
\hypersetup{
	colorlinks=true,
	citecolor=blue,
	urlcolor=blue,
	linkcolor=blue
}
\urlstyle{same}
\frenchspacing

\newcommand\partitle[1]{\textsc{#1}.}

@@

\title{On the reality of the quantum state once again}
\author{Gabriele Carcassi, Andrea Oldofredi, Christine A. Aidala}
\affiliation{Physics Department, University of Michigan, Ann Arbor, MI 48109}

@@

\begin{abstract}
We introduce an approach to the foundations of physics that is more in line with the foundations of mathematics. The idea is to examine current theories and find a set of starting physical assumptions that are sufficient to rederive them, which is the physical equivalent of the reverse mathematics program. This allows us to identify physical starting points (as opposed to mathematical ones), compare their deductive power, and find those that can better serve as a foundation for our theories.
\end{abstract}

@@

\section{Introduction}
%(mezza pagina)
%- Status of field and impact of HS
%- Claim that HS does not work for quantum epistemic cases, therefore invalidating all results

\section{Summary of the model}
%(una pagina)
%- Ontic/supplemented/epistemic

@@

Let us first characterize the extent of the discepancy. In HS, the space of epistemic states $E_{HS}$ is the set of probability distributions over pure states, which is an infinite dimensional space. This is a direct consequence of defining the epistemic states as a distribution $P(\lambda|P)$ that, for each preparation $P$, assigns a well defined probability to each ontological state $\lambda$.\footnote{TODO: HS seems to be very vague about probability and probability density.} In QM, mixed states $E_{Q}$ are in one-to-one correspondence with the interior of the Bloch sphere, a three dimensional manifold. Looking at the difference in dimension, the discrepancy is not minor.

Note that the space of pure states $\Lambda$ is the same in both cases, one state for each direction of spin. Also note that all epistemic states in both spaces can be reached by mixture of pure states. The difference is that, in the quantum case, we have an ``aliasing'': different mixtures will give the same mixed states. Since each classical probability distribution corresponds to a unique mixing of pure states, we have a map $\iota : E_{HS} \mapsto E_{Q}$ that allows us to go from HS epistemic states to density matrixes. This map is surjective but not injective. Therefore $E_{Q}$ can be seen as a set of equivalence classes of $E_{HS}$. The key question, then, is the following: is the map $\iota$ sufficiently well behaved that we can understand the epistemic content of $E_{Q}$ based on the epistemic content of $E_{HS}$? Or is it so hopelessy broken that there is no connection between the two?

@@

@@

\section*{Supplementary material}

TODO: do we need to bump anything here?

\bibliography{bibliography}

>>>>

%\documentclass[10pt,twocolumn, nofootinbib]{revtex4-2}
\documentclass[twocolumn,prl,floatfix,superscriptaddress]{revtex4-2}

\usepackage{amsthm, amsmath, amssymb, enumitem, graphicx, hyperref}
\usepackage{graphicx, subfigure}
\setlist{noitemsep}
\hypersetup{
	colorlinks=true,
	citecolor=blue,
	urlcolor=blue,
	linkcolor=blue
}
\urlstyle{same}

%\usepackage{graphicx, subfigure, amsfonts,amssymb,amsmath, array, gensymb,fontenc,color,lipsum}
%\documentclass[aps,twocolumn,showpacs,preprintnumbers]{revtex4}
%\usepackage{graphicx}  % Include figure files
%\usepackage{subfigure}
%\usepackage{multirow}
%\linespread{1.0}
%\usepackage{fancyhdr}
%\usepackage{longtable}
%\usepackage{parskip}
%\usepackage[T1]{fontenc}
%\usepackage{dcolumn}
%\usepackage{bm}        
%\usepackage{amsfonts}  
%\usepackage{amsmath}   
%\usepackage{amssymb}
%\usepackage{hyperref}
%\hypersetup{colorlinks= true, allcolors=blue}
%\setcitestyle{aysep={}}
%\setlength{\parindent}{11pt}
\frenchspacing

@@

\title{On the reality of the quantum state once again}
\author{Gabriele Carcass}
\affiliation{Physics Department, University of Michigan, Ann Arbor, MI 48109}
\author{Andrea Oldofredi}
\affiliation{Department of Philosophy, University of Lisbon, Portugal}
\author{Christine Aidala}
\affiliation{Physics Department, University of Michigan, Ann Arbor, MI 48109}
\vspace{2mm}

@@

\begin{abstract}
The ontological model framework introduced by Harrigan and Spekkens has been widely used to categorize quantum interpretations and as a basis for influential results such as the PBR theorem. In this paper we show that the model has a fundamental flaw: the epistemic structure it implicitly assumes does not follow the one dictated by quantum mechanics. Namely, the map between the epistemic states of the model and quantum density matrices is not even an order homomorphism in terms of the information entropy, consequently the epistemic content of mixed states is not mapped in a meaningful way. Moreover, by recasting the definitions of the model in terms of density matrices, we find that only $\psi$-epistemic interpretations are allowed by quantum mechanics, a result that is the opposite of the PBR theorem. We show that the cause of this mismatch is the effect of non-contextuality over mixed preparations that does not allow for a single decomposition of mixed states over pure states.
\end{abstract}

@@

\section{Introduction}

%- Status of field and impact of HS
%- Mention the PBR Theorem explicitly
%- Claim that HS does not work for quantum epistemic cases, therefore invalidating all results

\section{Summary of the Harrigan \& Spekkens Model}

\subsection{Ontological Models}

Harrigan and Spekkens (HS) provide a classification of ontological quantum models that essentially depends on the nature of the quantum state \cite{Harrigan:2010}. Such ontological models are defined by employing an operational setting where the primitive of descriptions consist in preparation protocols and measurements performed on quantum systems. More precisely, HS claim that preparations procedures are ``assumed to prepare a system with certain properties and a measurement procedure is assumed to reveal something about those properties'' (\cite{Harrigan:2010}, p.\ 128). In their approach the complete specification of the attributes of a given physical object is provided by $\lambda$, the ontic state of the system under scrutiny. Such operational models provide the probabilities $p(k|M, P)$ to obtain outcomes $k$ for some measurement $M$ performed on the prepared systems, given the set of preparation instructions $P$. Indeed, from a formal perspective every protocol $P$ is associated with a density operator $\rho$ on the relevant Hilbert space, and every measurement $M$ is associated with a POVM $\{ E_k\}$. As usual, each measurement result $k$ can be obtained with probability $p(k|M, P)=\textrm{Tr}(\rho E_k)$, the generalized Born rule. To this regard, HS underline that although an agent perfectly knows the preparation procedures prior the performance of a certain measurement, they may have incomplete knowledge of $\lambda$. This is means that an observer which has incomplete information about $\lambda$ assigns ``non-sharp'', i.e.\ overlapping probability distributions $p(\lambda| P)$ over $\Lambda$, the space of ontic states.

We can eventually provide the formal definition of ontological models as follows (\cite{Harrigan:2010}, p.\ 128): ``An ontological model of operational quantum theory posits an ontic state space $\Lambda$ and prescribes a probability distribution over $\Lambda$ for every preparation procedure $P$, denoted $p(\lambda| P)$, and a probability distribution over the different outcomes $k$ of a measurement $M$ for every ontic state $\lambda \in \Lambda$, denoted $p(k|\lambda, M)$. Finally, for all $P$ and $M$, it must satisfy
\begin{align*}
\int d\lambda p(k|\lambda, M) p(\lambda| P)= \textrm{tr}(\rho E_k),
\end{align*}
\noindent where $rho$ is the the density operator associated with $P$ and $E_k$ is the POVM element associated with outcome $k$ of $M$''.

\subsection{$\psi-$ontic and $\psi-$epistemic models}

Let us now introduce the first categorization of ontological models. HS divide them in $\psi-$\emph{complete} and $\psi-$\emph{incomplete}: in the former case, the quantum state encodes every information about the physical system at hand, as in the case of standard Quantum Mechanics (QM). Alternatively stated, in $\psi$-complete models there is a one-to-one relation between pure quantum states and the ontic state of physical items they represent. Formally, this is translated into the following definition (\cite{Harrigan:2010}, p.\ 131): ``An ontological model is $\psi-$complete if the ontic state space $\Lambda$ is isomorphic to the projective Hilbert space $\mathcal{PH}$ (the space of rays of Hilbert space) and if every preparation procedure $P_{\psi}$ associated in quantum theory with a given ray $\psi$ is associated in the ontological model with a Dirac delta function centered at the ontic
state $\lambda_{\psi}$ that is isomorphic to $\psi$, $p(\lambda|P)=\delta(\lambda-\lambda_{\psi})$''.
In the latter case, $\psi$ alone is not sufficient to provide a complete description of reality. Models in which $\psi$ is supplemented with additional/hidden variables are defined by HS as $\psi-$supplemented (well-known examples are Bohm pilot-wave theory \cite{Bohm:1952aa} and Nelson mechanics \cite{Nelson:1966aa}).

In the second place, the authors propose another distinction concerning the nature of quantum states, classifying $\psi$ either as \emph{ontic} or \emph{epistemic}. In a nutshell, a model is said to be $\psi-$ontic if ``for any pair of preparation procedures, $P_{\psi}$ and $P_{\phi}$, associated with distinct quantum states $\psi$ and $\phi$, we have $p(\lambda | P_{\psi})p(\lambda|P_{\phi})=0$ for all $\lambda$'' (\cite{Harrigan:2010}, pp.\ 131-132). Consequently, (i) in $\psi-$ontic models the ontic state $\lambda$ can be consistently described only by a unique pure quantum state $\psi$, and (ii) observers' epistemic states associated with different quantum states are non-overlapping in $\Lambda$, i.e.\ they correspond to disjoint regions of the space of ontic states. Alternatively, in $\psi-$ontic models different pure quantum states correspond to disjoint probability distributions $\mu$ over $\Lambda$ (cf.\ Figure 1a).

If a model in not $\psi-$ontic, then it is $\psi-$epistemic. A model is defined $\psi-$epistemic if $\lambda$ can be described by more than one quantum state. Therefore, in such epistemic models there are quantum states that correspond to overlapping probability distributions over $\Lambda$ (cf.\ Figure 1b).
This fact implies that agents' epistemic states overlap, i.e.\ there exist preparation procedures $P_{\psi}, P_{\phi}$ and a $\lambda\in\Lambda$ such that $p(\lambda | P_{\psi})p(\lambda|P_{\phi})\neq0$. In turn, this entails that the ontic state $\lambda$ can be consistently represented by both quantum states $\psi$ and $\phi$: ``[i]n a $\psi-$epistemic model, multiple distinct quantum states are consistent with the same state of reality---the ontic state $\lambda$ does not encode $\psi$'' (\emph{ibid.}, p.\ 132). This explains why the authors claim that in $\psi-$epistemic models quantum states refer to observers' incomplete knowledge of reality, and not to reality itself.

\begin{figure}
\includegraphics[scale=.7]{ontic}
\caption{\footnotesize{Harrigan and Spekkens distinction between $\psi-$ontic (a) and $\psi-$epistemic (b) ontological models.}}
\end{figure}

Now we can combine both distinctions in order to classify different ontological models. As we have seen, in $\psi-$complete models there is a one-to-one relation between $\lambda$ and its description provided by $\psi$. Consequently, if one knows the quantum state $\psi$ then one has a complete knowledge of the ontic state $\lambda$ of the system under consideration. In this models a variation of $\psi$ implies a variation of the ontic state, therefore $\psi-$complete models are $\psi-$ontic. If ontological models are $\psi-$incomplete, then they may be either $\psi-$supplemented or $\psi-$epistemic. In the former case the description of a physical system provided by $\psi$ is incomplete and is supplemented with additional (or hidden) variables whose value is generally unknown. Notably, also the class of $\psi-$supplemented models is $\psi-$ontic. Finally, in $\psi-$epistemic models $\psi$ represents an agent's incomplete knowledge of reality (of $\lambda$), and not reality itself; in this case a variation of $\psi$ does not necessarily entail a variation of $\lambda$.

From these distinctions some conclusions can be inferred. In the first place, models which are simultaneously $\psi-$complete \emph{and} $\psi-$epistemic cannot exist. Therefore, if a model is $\psi-$complete, it must be $\psi-$ontic (cf.\ Lemma 6, \cite{Harrigan:2010}, p.\ 133). Alternatively, if a model is $\psi-$incomplete, then it can either be $\psi-$ontic, as in the case of $\psi-$supplemented models, or $\psi-$epistemic. If a model is $\psi-$epistemic, then it cannot be $\psi-$ontic, since it does not describe any underlying physical reality, but only the agents' knowledge of it.

@@

Let us first characterize the extent of the discrepancy. In HS, the space of epistemic states $E_{HS}$ is the set of probability distributions over pure states, which is an infinite dimensional space. This is a direct consequence of defining the epistemic states as a distribution $P(\lambda|P)$ that, for each preparation $P$, assigns a well defined probability to each ontological state $\lambda$.\footnote{TODO: HS seems to be very vague about probability and probability density.} In QM, mixed states $E_{Q}$ are in one-to-one correspondence with the interior of the Bloch sphere, a three dimensional manifold. Looking at the difference in dimension, the discrepancy is not minor.

Note that the space of pure states $\Lambda$ is the same in both cases, one state for each direction of spin. Also note that all epistemic states in both spaces can be reached by mixture of pure states. The difference is that, in the quantum case, we have an ``aliasing'': different mixtures will give the same mixed states. Since each classical probability distribution corresponds to a unique mixing of pure states, we have a map $\iota : E_{HS} \mapsto E_{Q}$ that allows us to go from HS epistemic states to density matrixes. This map is surjective but not injective. Therefore $E_{Q}$ can be seen as a set of equivalence classes of $E_{HS}$. The key question, then, is the following: is the map $\iota$ sufficiently well behaved that we can understand the epistemic content of $E_{Q}$ based on the epistemic content of $E_{HS}$? Or is it so hopelessly broken that there is no connection between the two?

@@

\begin{figure}
\includegraphics[scale=.4]{fig2}
\caption{\footnotesize{}}
\end{figure}

@@

%\section*{Supplementary material}
%
%TODO: do we need to bump anything here?

\bibliographystyle{apalike}
\bibliography{PhDthesis}
